Sampling requirements to detect ecosystem change

### 2020-02-27 R/manuscript_draft.Rmd
contribution to writing

Title: Sampling requirements to detect ecosystem shifts

@@

Environmental monitoring is a key piece of managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, soil or ice cores are taken in the field and later processed in a laboratory setting. Standard approaches either involve either equally-spaced or random sampling throughout each core. We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change. We test these ideas with a case study of paleoecological plankton data extracted from lake sediment cores. We demonstrate that standard approaches to sample processing are inefficient and costly. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools.

@@

```{r load_packages,echo=F,warning=F,message=F}
if (!require("pacman",character.only = TRUE))
  {
    install.packages("pacman",dep=TRUE)
    if(!require("pacman",character.only = TRUE)) stop("Package not found")
  }
library("pacman")

@@

Environmental monitoring is one of the core components to modern research and management (citations). Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (Lovett et al, 2007). Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010). However, environmental monitoring is still often expensive and time-consuming (citations). Therefore, monitoring programs need to be designed in such a way to address the question of interest. Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation). Monitoring programs also need to be designed to be cost-effective given limited budgets (citations).

Methods to detect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html

- Historical data and ability to choose how much to sample

@@

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie). Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA (citation). Likewise, video-graphical approaches can record snapshots of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed at every minute? As long as processing samples is expensive, these trade-offs will remain.

@@

The random sampling approach involves taking a set number of random points throughout the time series (Fig. \ref{fig:conceptual}c). In the context of sediment cores, this would mean analyzing community composition as random spots along the core. We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint. Regular sampling is commonly used (citations) and requires that a set number of samples be taken at regular intervals (Fig. \ref{fig:conceptual}d). Lastly, iterative sampling involves first taking a set number of samples and then alliteratively adding samples until a pre-determined level of precision is achieved (Fig. \ref{fig:conceptual}e-g).

>>>>

Title: Sampling requirements to detect ecosystem shifts
(lame title) Getting closer to the point: sampling requirements to detect ecosystem shifts.
Recommendation for temporal sampling

@@

Environmental monitoring is a key piece of managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be taken in the field in less than an hour, but will be processed in a laboratory setting for anytime between a week and several months, depending on how much information is extracted. Sub-sampling allows to reduce the processing time while still giving precious insight. Standard sub-sampling approaches often involve equally-spaced sampling. We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage, which trajectory was reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are inefficient, hence costly. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.

@@

```{r load_packages,echo=F,warning=F,message=F}
if (!require("pacman",character.only = TRUE))
  {
    install.packages("pacman",dep=TRUE)
    if(!require("pacman",character.only = TRUE)) stop("Package not found")
  }

@@

Environmental monitoring is one of the core components to modern research and management (citations). Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (Lovett et al, 2007). Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010). However, environmental monitoring is still often expensive and time-consuming, especially when further analyses are needed following sample collection (citations). In Ecology and Genetics especially, there has always been a disparity between the amount of data that can be aquired and stored, and the processing methods that could allow to extract the most information from that data. \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths. With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at. If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}

Therefore, monitoring programs need to be designed in such a way to address the question of interest, without wasting resources. Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation). Monitoring programs also need to be designed to be cost-effective given limited budgets (citations).

<!-- 
Methods to detect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html

- Historical data and ability to choose how much to sample
-->

@@

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, the field of paleolimnology paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie). Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA (citation). Likewise, video-graphical approaches can record snapshots of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed at every minute? As long as processing samples is expensive, these trade-offs will remain.

@@

The random sampling approach involves taking a set number of random points throughout the time series (Fig. \ref{fig:conceptual}c). In the context of sediment cores, this would mean analyzing community composition as random location along the core. Random sampling is recommanded in designs aiming at quantifying the average size of a population (spatial approach) (Nad’o and Kaňuch, 2018). We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint. Regular sampling is commonly used (citations) and requires that a set number of samples be taken at regular intervals (Fig. \ref{fig:conceptual}d). Lastly, iterative sampling involves first taking a set number of samples and then alliteratively adding samples until a pre-determined level of precision is achieved (Fig. \ref{fig:conceptual}e-g).
